The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?

### 2023-02-03 data-testing.qmd
Update comments and descriptions; add .pdf files to gitignore

# Test that the actual performance data is greater than or equal to 0%
raw_response_data$performance_in_percentage |> min() >= 0.00

```

@@

# Test that the incident type contains is comprised of characters
raw_incident_data$incident_type |> class() == "character"

```

>>>>

# Test that the actual performance data is greater than or equal to 0%
raw_response_data$performance_in_percentage |> min() >= 0.00
```

@@

# Test that the incident type contains is comprised of characters
raw_incident_data$incident_type |> class() == "character"
```

### 2023-02-03 inputs/data/data_import.qmd
Update comments and descriptions; add .pdf files to gitignore

```{r}
# Define required libraries
library(opendatatoronto)
library(tidyverse)
library(dplyr)

# Fetch Open Data Toronto's "Land Ambulance Response Time Standard" package: https://open.toronto.ca/dataset/land-ambulance-response-time-standard 
raw_response_data <-
  list_package_resources("8b068db1-95b8-4ea9-b942-8a7ce3a5fbd6") |>
  filter(name == "Response Time (2013 - 2021)") |>
  get_resource()

# Write data to a .csv file
write_csv(
  x = raw_response_data$`Land Ambulance Services`,
  file = "raw_response_data.csv"
)

# Fetch Open Data Toronto's "Pre-Hospital Emergency Care Performance Metrics" package:  https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
raw_emergent_care_data <-
  list_package_resources("174ace54-bcc7-4885-85c1-76094ab4a0ca") |>
  filter(name == "Pre-hospital emergency care performance metrics 2014-2021") |>
  get_resource() |>
  filter(`Pre-Hospital Emergency Care Type` == "Emergency Request for Service")

# Write data to a .csv file
write_csv(
  x = raw_emergent_care_data,
  file = "raw_emergent_care_data.csv"
)

# Fetch Open Data Toronto's "Paramedic Services Incident Data" package: https://open.toronto.ca/dataset/paramedic-services-incident-data
raw_incident_data <-
  list_package_resources("c21f3bd1-e016-4469-abf5-c58bb8e8b5ce") |>
  filter(name == "paramedic-services-incident-data-2017-2021") |>
  get_resource()

>>>>

<!--
#### Preamble ####
# Purpose: Import data from the Toronto Paramedic Service via Open Data Toronto
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses
-->
---
title: "Import data from the Toronto Paramedic Service via Open Data Toronto"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Workspace set-up ####
library(opendatatoronto)
library(tidyverse)
library(dplyr)
```

```{r}
# Fetch Open Data Toronto's "Land Ambulance Response Time Standard" package: https://open.toronto.ca/dataset/land-ambulance-response-time-standard 
raw_response_data <-
  list_package_resources("8b068db1-95b8-4ea9-b942-8a7ce3a5fbd6") |>
  filter(name == "Response Time (2013 - 2021)") |>
  get_resource()

# Write data to a .csv file
write_csv(
  x = raw_response_data$`Land Ambulance Services`,
  file = "raw_response_data.csv"
)
```

```{r}
# Fetch Open Data Toronto's "Pre-Hospital Emergency Care Performance Metrics" package:  https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
raw_emergent_care_data <-
  list_package_resources("174ace54-bcc7-4885-85c1-76094ab4a0ca") |>
  filter(name == "Pre-hospital emergency care performance metrics 2014-2021") |>
  get_resource() |>
  filter(`Pre-Hospital Emergency Care Type` == "Emergency Request for Service")

# Write data to a .csv file
write_csv(
  x = raw_emergent_care_data,
  file = "raw_emergent_care_data.csv"
)
```

```{r}
# Fetch Open Data Toronto's "Paramedic Services Incident Data" package: https://open.toronto.ca/dataset/paramedic-services-incident-data
raw_incident_data <-
  list_package_resources("c21f3bd1-e016-4469-abf5-c58bb8e8b5ce") |>
  filter(name == "paramedic-services-incident-data-2017-2021") |>
  get_resource()

### 2023-02-03 outputs/paper/analysis.qmd
Update comments and descriptions; add .pdf files to gitignore

---
title: "The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?"
format: pdf
author: "Sebastian Rodriguez^[University of Toronto, Faculty of Information]"
abstract: "The Toronto Paramedic Services provides all pre-hospital care to the city’s inhabitants. Recent criticism of the service called into question its performance, highlighting staff shortages and institutional challenges that the city's paramedics face. This paper investigates the Toronto Paramedic Services’ efficacy by comparing its emergency response times to the standards set by provincial and municipal governments. It reveals that the Toronto's paramedics meet and often exceed response time standards, and provided stroke and STEMI heart attack patients with a chance of survival greater than the provincial average." 
date: today
date-format: long
editor: visual
toc: TRUE
number-sections: TRUE
bibliography: references.bib
thanks: "Code and data supporting this analysis is available at: https://github.com/seb646/toronto-paramedic-responses"
---

## Introduction

A city's emergency medical services (EMS) are integral to its health care system. The paramedics and emergency medical technicians working for the service provide pre-hospital assessment, stabilisation, and care to people needing medical assistance. These professionals respond to every call from the public for aid, from broken bones and heart attacks to car accidents and house fires. Current emergency medical services operate on the principle of the "golden hour," which is the 60 minutes between an initial injury and arrival at a trauma center when a patient has the highest statistical chance of survival [@cite_golden_hour]. This "golden hour" is the foundation of all emergency medical responses and typically requires medical personnel to arrive on the scene of an injury in under 10 minutes [@cite_golden_hour_10_mins].

## Paramedic Response Data

```{r}
#| messages: false
#| echo: false
#| warning: false
# Define required libraries 
library(lubridate)
library(tidyverse)
library(dplyr)
library(knitr)
library(janitor)
library(scales)
library(RColorBrewer)
library(ggplot2)
library(kableExtra)
library(here)
library(readr)

@@

### Scope of Responsibility

Before investigating the response times and efficacy of the Toronto Paramedic Services, it is important to understand the service's breadth of responsibility. With a population of nearly 3 million people, Toronto requires a medical response to roughly 300,000 emergency calls every year. As the city's population grows, so too does the need for medical assistance. There was a steady increase of approximately 10,000 emergency medical requests per year from 2014 to 2019, as outlined by the graph below (@fig-1). This trend ended briefly in 2020 when the COVID-19 pandemic hit Toronto, placing an incredible strain on the city's healthcare system. Due to a concern about COVID-19 infection, many people delayed or avoided medical care throughout the pandemic [@cite_avoiding_hospitals].

@@

\newpage

@@

# Render table
clean_incident_data |>
  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
  add_header_above(
    c(" ", "Year Totals" = 2, "Dispatch Averages" = 2), 
    align = "c",
    bold=TRUE
  ) |>
  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
  row_spec(0,bold=TRUE) |>
  column_spec(1, width = "13em") 
```

\newpage

### Mandated Response Times

In Ontario, the *Ambulance Act* defines the rules governing paramedics' response times. The act defines response times as the difference between the time an emergency service dispatcher receives a call and the on-scene arrival of either an ambulance or a person equipped with a defibrillator [@cite_ambulance_act]. The *Ambulance Act* mandates varying response times based on the severity of a patient's injury. To define these response times and categorize patients based on their injuries, paramedics use the Canadian Triage Acuity Scale (CTAS). CTAS has 5 levels of severity, with level 1 representing life-threatening injuries and level 5 representing non-urgent injuries.

@@

### Actual Response Times

The Toronto Paramedic Services has an excellent track record of achieving, and often exceeding, the percentage of calls receiving mandated response times. The graphs below (@fig-4) chart the percentage of incidents served by the mandated response times versus the city's set performance plan. To place these graphs in perspective, the City of Toronto's paramedic service averaged a response time of 6 minutes and 58 seconds for "emergency calls" in 2021 [@cite_ambulance_program].

@@

## Discussion

The data used for analysis in this paper was collected and published by the Toronto Paramedic Service. As such, the data is open to institutional bias. The Toronto Paramedic Services determines what data is collected and the methods used for collection. It also has specific goals for data collection: in the case of response times, they must collect the data so they can report their performance to Ontario's Ministry of Health. These goals can impact the data and cause the service to omit significant data points, such as the lack of a measured, case-by-case response time. The data also has other limitations, such as the lack of metrics after 2021.

@@

**Land Acknowledgement**

We wish to acknowledge the land on which this data was collected and analysed. For thousands of years, it has been the traditional land of the Huron-Wendat, the Seneca, and the Mississaugas of the Credit. Today, this meeting place is still the home to many Indigenous people from across Turtle Island and we are grateful to have the opportunity to work on this land.

@@

**Data Acknowledgement**

The data used in this paper was obtained through Open Data Toronto, an initiative to provide openly accessible data about the city [@cite_about_opendatatoronto]. This paper uses three data sets from Open Data Toronto, each provided by the Toronto Paramedic Services:

@@

**Resources Acknowledgment**

The primary tool used to analyse data in this paper is R, an open-source statistical programming language [@cite_tool_r]. The paper also uses a number of R packages, including: dplyr [@cite_tool_dplyr], ggplot2 [@cite_tool_ggplot2], here [@cite_tool_here], janitor [@cite_tool_janitor], kableExtra [@cite_tool_kableextra], knitr [@cite_tool_knitr], lubridate [@cite_tool_lubridate], opendatatoronto [@cite_tool_opendatatoronto], readr [@cite_tool_readr], RColorBrewer [@cite_tool_rcolorbrewer], scales [@cite_tool_scales], and tidyverse [@cite_tool_tidyverse]. \newpage

## References
<!-- Span required to show references as a numbered section -->

>>>>

<!--
#### Preamble ####
# Title: The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses
-->
---
title: "The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?"
format: pdf
author: "Sebastian Rodriguez^[University of Toronto, Faculty of Information]"
abstract: "The Toronto Paramedic Services provides all pre-hospital care to the city’s inhabitants. Recent criticism of the service called into question its performance, highlighting staff shortages and institutional challenges that the city's paramedics face. This paper investigates the Toronto Paramedic Services’ efficacy by comparing its emergency response times to the standards set by provincial and municipal governments. It reveals that the Toronto's paramedics meet and often exceed response time standards, and provided stroke and STEMI heart attack patients with a chance of survival greater than the provincial average." 
date: today
date-format: long
editor: visual
toc: TRUE
number-sections: TRUE
bibliography: references.bib
thanks: "Code and data supporting this analysis is available at: https://github.com/seb646/toronto-paramedic-responses"
---
```{r}
#| messages: false
#| echo: false
#| warning: false
#### Workspace set-up ####
# Define required libraries 
library(lubridate)
library(tidyverse)
library(dplyr)
library(knitr)
library(janitor)
library(scales)
library(RColorBrewer)
library(ggplot2)
library(kableExtra)
library(here)
library(readr)

@@

## Introduction
A city's emergency medical services (EMS) are integral to its health care system. The paramedics and emergency medical technicians working for the service provide pre-hospital assessment, stabilisation, and care to people needing medical assistance. These professionals respond to every call from the public for aid, from broken bones and heart attacks to car accidents and house fires. Current emergency medical services operate on the principle of the "golden hour," which is the 60 minutes between an initial injury and arrival at a trauma center when a patient has the highest statistical chance of survival [@cite_golden_hour]. This "golden hour" is the foundation of all emergency medical responses and typically requires medical personnel to arrive on the scene of an injury in under 10 minutes [@cite_golden_hour_10_mins].

## Paramedic Response Data
### Scope of Responsibility
Before investigating the response times and efficacy of the Toronto Paramedic Services, it is important to understand the service's breadth of responsibility. With a population of nearly 3 million people, Toronto requires a medical response to roughly 300,000 emergency calls every year. As the city's population grows, so too does the need for medical assistance. There was a steady increase of approximately 10,000 emergency medical requests per year from 2014 to 2019, as outlined by the graph below (@fig-1). This trend ended briefly in 2020 when the COVID-19 pandemic hit Toronto, placing an incredible strain on the city's healthcare system. Due to a concern about COVID-19 infection, many people delayed or avoided medical care throughout the pandemic [@cite_avoiding_hospitals].

@@

```
\newpage

@@

# Render table
clean_incident_data |>
  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
  add_header_above(
    c(" ", "Year Totals" = 2, "Dispatch Averages" = 2), 
    align = "c",
    bold=TRUE
  ) |>
  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
  row_spec(0,bold=TRUE) |>
  column_spec(1, width = "13em") 
```
\newpage

### Mandated Response Times
In Ontario, the *Ambulance Act* defines the rules governing paramedics' response times. The act defines response times as the difference between the time an emergency service dispatcher receives a call and the on-scene arrival of either an ambulance or a person equipped with a defibrillator [@cite_ambulance_act]. The *Ambulance Act* mandates varying response times based on the severity of a patient's injury. To define these response times and categorize patients based on their injuries, paramedics use the Canadian Triage Acuity Scale (CTAS). CTAS has 5 levels of severity, with level 1 representing life-threatening injuries and level 5 representing non-urgent injuries.

@@

### Actual Response Times
The Toronto Paramedic Services has an excellent track record of achieving, and often exceeding, the percentage of calls receiving mandated response times. The graphs below (@fig-4) chart the percentage of incidents served by the mandated response times versus the city's set performance plan. To place these graphs in perspective, the City of Toronto's paramedic service averaged a response time of 6 minutes and 58 seconds for "emergency calls" in 2021 [@cite_ambulance_program].

@@

## Discussion
The data used for analysis in this paper was collected and published by the Toronto Paramedic Service. As such, the data is open to institutional bias. The Toronto Paramedic Services determines what data is collected and the methods used for collection. It also has specific goals for data collection: in the case of response times, they must collect the data so they can report their performance to Ontario's Ministry of Health. These goals can impact the data and cause the service to omit significant data points, such as the lack of a measured, case-by-case response time. The data also has other limitations, such as the lack of metrics after 2021.

@@

**Land Acknowledgement**
We wish to acknowledge the land on which this data was collected and analysed. For thousands of years, it has been the traditional land of the Huron-Wendat, the Seneca, and the Mississaugas of the Credit. Today, this meeting place is still the home to many Indigenous people from across Turtle Island and we are grateful to have the opportunity to work on this land.

@@

**Data Acknowledgement**
The data used in this paper was obtained through Open Data Toronto, an initiative to provide openly accessible data about the city [@cite_about_opendatatoronto]. This paper uses three data sets from Open Data Toronto, each provided by the Toronto Paramedic Services:

@@

**Resources Acknowledgment**
The primary tool used to analyse data in this paper is R, an open-source statistical programming language [@cite_tool_r]. The paper also uses a number of R packages, including: dplyr [@cite_tool_dplyr], ggplot2 [@cite_tool_ggplot2], here [@cite_tool_here], janitor [@cite_tool_janitor], kableExtra [@cite_tool_kableextra], knitr [@cite_tool_knitr], lubridate [@cite_tool_lubridate], opendatatoronto [@cite_tool_opendatatoronto], readr [@cite_tool_readr], RColorBrewer [@cite_tool_rcolorbrewer], scales [@cite_tool_scales], and tidyverse [@cite_tool_tidyverse]. \newpage

## References
<!-- Comment required to show references as a numbered section -->
